Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis

### 2021-06-10 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Submission to Geographical Analysis

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={The importance of reproducibility in geographical research: the case of population density and the spread of COVID-19},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{The importance of reproducibility in geographical research: the
case of population density and the spread of COVID-19}
    \author[University]{Author\corref{1}}
   \ead{author@institution.edu} 
      \address[University]{Department, Street, City, State ZIP}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  The emergence of the novel SARS-CoV-2 coronavirus and the global
  COVID-19 pandemic has led to explosive growth in scientific research.
  Of interest has been the associations between population density and
  the spread of the pandemic. In this paper, population density and the
  basic reproductive number of SARS-CoV-2 are examined in an example of
  reproducible research. Given the high stakes of the situation, it is
  essential that scientific activities, on which good policy depends,
  are as transparent and reproducible as possible. Reproducibility is
  key for the efficient operation of the self-correction mechanisms of
  science. Transparency and openness means that the same problem can,
  with relatively modest efforts, be examined by independent researchers
  who can verify findings, and bring to bear different perspectives,
  approaches, and methods---sometimes with consequential changes in the
  conclusions, as the empirical example of the spread of COVID-19 in the
  US shows.
  \end{abstract}

@@

Given the heavy human and economic cost of the pandemic, there has been
a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has termed this
deluge of publications a ``carnage of substandard research'' (Bramstedt,
2020). Part of the challenge of maintaining quality standards in
published research is that, despite an abundance of recommendations and
guidelines (Broggini et al., 2017; Brunsdon and Comber, 2020; Ince et
al., 2012; Ioannidis et al., 2014), in practice reproducibility has
remained a lofty and somewhat aspirational goal (Konkol et al., 2019;
Konkol and Kray, 2019). As reported in the literature, only a woefully
small proportion of published research was actually reproducible before
the pandemic (Iqbal et al., 2016; Stodden et al., 2018), and the
situation does not appear to have changed substantially since (Gustot,
2020; Sumner et al., 2020).

@@

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic (Florida et al., 2020); the implications are the topic of
ongoing research (Sharifi and Khavarian-Garmsir, 2020). The fact that
the worst of the pandemic was initially felt in dense population centers
such as Wuhan, Milan, Madrid, and New York, brought a torrent of
research into the associations between density and the spread of the
pandemic. Some important questions hang on the results of these research
efforts. For example, are lower density regions safer from the pandemic?
Are de-densification policies warranted, even if just in the short term?
And in the longer term, will the risks of life in high density regions
presage a flight from cities? What are the implications of the pandemic
for future urban planning and practice? Over the past year, numerous
papers have sought to throw light into the underlying issue of density
and the pandemic; nonetheless the results, as will be detailed next,
remain mixed. Further, to complicate matters, precious few of these
studies appear to be sufficiently open to support independent
verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. To this end, a recent study by Sy et al. (2021) is chosen as an
example of reproducible research. The objective is not to malign the
analysis of these researchers, but rather to demonstrate the value of
openness to allow for independent verification and further analysis.
Open data and open code mean that an independent researcher can, with
only modest efforts, not only verify the findings reported, but also
examine the same data from a perspective which may not have been
available to the original researchers due to differences in disciplinary
perspectives, methodological traditions, and/or training, among other
possible factors. The example, which shows consequential changes in the
conclusions reached by different analyses, should serve as a call to
researchers to redouble their efforts to increase transparency and
reproducibility in research. This paper, in addition, aims to show how
data can be packaged in well-documented, shareable units, and code can
be embedded into self-contained documents suitable for review and
independent verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, will be available in a public
repository\footnote{For peer-review purposes, the data package and code
  are currently in an anonymous Drive folder:
  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.

@@

As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
mathematical models of contagion are valid at small-to-medium spatial
scales (and presumably, small temporal scales too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and different time scales. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers-however, a person can choose to
(or be forced to as a matter of policy) not go to a restaurant in the
first place. Nonetheless, the idea that high density correlates with
high transmission is so seemingly reasonable that it is often taken for
granted even at larger scales (e.g., Cruz et al., 2020; Micallef et al.,
2020). At larger scales, however, there exists the possibility of
behavioral adaptations, which are difficult to capture in the
mechanistic framework of differential equations (or can be missing in
agent-based models, e.g., Gomez et al., 2021); these adaptations, in
fact, can be a key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk (Noland, 1995; Phillips et
al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
al. (Chauhan et al., 2021) have found that perception of risks in the US
varies between rural, suburban, and urban residents, with rural
residents in general expressing less concern about the virus. It is
possible that people who listened to the message of leaders saying that
they were safe because of low density may not have taken adequate
precautions against the virus. People in dense places who could more
directly observe the impact of the pandemic may have become overly
cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
mechanism (i.e., greater compliance with social distancing in denser
regions) to explain the results of their analyses. The evidence
available does indeed show that there were important changes in behavior
with respect to mobility during the pandemic (Harris and Branion-Calles,
2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
in place orders may have had greater buy-in from the public in higher
density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
and the associated behavior may have persisted beyond the duration of
official social-distancing policies (Praharaj et al., 2020). In
addition, there is evidence that changes in mobility correlated with the
trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. (2020) who did not find evidence of
significant correlation between population density and confirmed number
of cases per day \emph{in conditions of lockdown} in China. This finding
echoes the results of Paez et al. (2020), who in their study of
provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, (2020) found zero or negative
associations between population density and infection numbers/deaths by
country. Fielding-Miller et al. (2020) contrast their finding about
rural counties with a negative relationship between COVID-19 deaths and
population density in urban counties in the US. For their part, in their
investigation of doubling time, White and Hébert-Dufresne (2020)
identified a negative and significant correlation between population
density and doubling time in US states. Likewise, (2021) fond a small
negative (and significant) association between population density and
COVID-19 morbidity in districts in Tehran. Finally, two of the most
complete studies in the US {[}by Hamidi et al.; (2020a); (2020b){]} used
an extensive set of controls to find negative and significant
correlations between density and COVID-19 cases and fatalities at the
level of counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
of a researcher with a somewhat different toolbox.

@@

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. For their multivariate
analysis, SWN used mixed linear models. This is a reasonable modelling
choice: \(R_0\) is an interval-ratio variable that is suitably modeled
using linear regression; further, as SWN note there is a likelihood that
the process in not independent ``among counties within each state,
potentially due to variable resource allocation and differing health
systems across states'' (p.~3). A mixed linear model accounts for this
by introducing random components (in the case of SWN, random intercepts
at the state level). SWN estimated various models with different
combinations of variables, including median household income and
prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SWN conclude that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

One salient aspect of the analysis in SWN is that the basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made modelling decisions, in this case basing their
analysis only on counties with valid observations. A modeler with
expertise in different methods would likely ask some of the following
questions on reading SWN's paper: how were missing counties treated?
What are the implications of the spatial sampling framework used in the
analysis? Is it possible to spatially interpolate the missing
observations? These questions are relevant and their implications
important. Fortunately, SWN are an example of a reasonably open,
reproducible research product: their paper is accompanied by (most of)
the data and (most of) the code used in the analysis. This means that an
independent expert can, with only a moderate investment of time and
effort, reproduce the results in the paper, as well as ask additional
questions.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
2020). But in many other cases the data are publicly available, which in
fact has commonly been the case with population-level COVID-19
information. Typically the provenance of the data is documented, but in
numerous studies the data themselves are not shared (Amadu et al., 2021;
Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
whether a graduate student or a seasoned scientist, collecting,
organizing, and preparing data for a project can take a substantial
amount of time. Pointing to the sources of data, even when these sources
are public, is a small step towards reproducibility-but only a very
small one. Faced with the prospect of having to recreate a data set from
raw sources is probably sufficient to dissuade all but the most
dedicated (or stubborn) researcher from independent verification. This
is true even if part of the data are shared (e.g., Wong and Li, 2020).
In other cases, data are shared, but the processes followed in the
preparation of the data are not fully documented (Ahmad et al., 2020;
Skórka et al., 2020). These processes matter, as shown by the errors in
the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis (Noury et al., 2021; Pequeno et
al., 2020; Wang et al., 2021). Making code available only ``on demand''
(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
journals offer the facility to share supplemental materials online. Then
there are those papers that more closely comply with reproducibility
standards, and share well-documented processes and data, as well as the
code used in any analyses reported (Feyman et al., 2020; Paez et al.,
2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
2020).

In the following sections, the analysis of RWN is reproduced, some
relevant questions from the perspective of an independent researcher are
asked, and the data are reanalyzed.

@@

Table \ref{tab:swn-results} reproduces the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SWN). It is possible to verify that the results match, with only the
minor (and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is one percent units, whereas in SWN it was ten
percent units). The mixed linear model gives random intercepts (i.e.,
the intercept is a random variable), and the standard deviation is
reported in the fourth row of Table \ref{tab:swn-results}. It is useful
to map the random intercepts: as seen in Figure
\ref{fig:random-terms-map}, other things being equal, counties in Texas
tend to have somewhat lower values of \(R_0\) (i.e., a negative random
intercept), whereas counties in South Dakota tend to have higher values
of \(R_0\). The key of the analysis, after extensive sensitivity
analysis, is a robust finding that population density has a positive
association with the basic reproductive number. But does it?

@@

The preceding section shows that thanks to the availability of code and
data, it is possible to verify the results reported by SWN. As noted
earlier, though, an independent researcher might have wondered about the
implications of the spatial sampling procedure used by SWN. The decision
to use a sample of counties with reliable basic reproductive numbers,
although apparently sensible, results in a non-random spatial sampling
scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
the impression that many counties without reliable values of \(R_0\) are
in more rural, less dense parts of the United States. This impression is
reinforced when we overlay the boundaries of urban areas with population
greater than 50,000 on the counties with valid values of \(R_0\) (see
Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
accurately computed in many counties without large urban areas does not
mean that there was no transmission of the virus: it simply means that
we do not know with precision whether that was the case. The low number
of cases may be related to low population and/or low population density.
This is intriguing, to say the least: by excluding cases based on the
ability to calculate \(R_0\) we are potentially \emph{censoring} the
sample in a non-random way.

@@

A problematic issue with non-random sample selection is that parameter
estimates can become unreliable, and numerous techniques have been
developed over time to address this. A model useful for sample selection
problems is Heckman's selection model (see Maddala, 1983). The selection
model is in fact a system of two equations, as follows:

\[
\begin{array}{c}
y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
\end{array}
\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
\(\epsilon_i^S\) and \(\epsilon_i^O\)) The first equation is designed to
model the \emph{probability} of sampling, and the second equation the
outcome of interest (say \(R_0\)). The random terms are jointly
distributed and correlated with parameter \(\rho\).

@@

A sample selection model is estimated using the same selection of
variables as SWN Model 3. This is Sample Selection Model 1 in Table
\ref{tab:selection-results}. The first thing to notice about this model
is that the sample selection process and the outcome are not independent
(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
that the probability of a county to be in the sample increases with
population density (but at a decreasing rate due to the
log-transformation), when travel by private modes is more prevalent, and
as median household income in the county is higher. This is in line with
the impression left by Figure \ref{fig:urban-areas-map} that counties
with reliable values of \(R_0\) tended to be those with larger urban
centers. Once that the selection probabilities are accounted for in the
model, several things happen with the outcomes model. First, the
coefficient for population density is still positive, but the magnitude
changes: in effect, it appears that the effect of density is more
pronounced than what SWN Model 3 indicated. The coefficient for percent
of private transportation changes signs. And the coefficient for median
household income is now significant.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{setspace}\onehalfspacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Reproducibility of research during COVID-19: examining the case
of population density and the basic reproductive rate from the
perspective of spatial analysis}
    \author[University]{Author\corref{1}}
   \ead{author@institution.edu} 
      \address[University]{Department, Street, City, State ZIP}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  The emergence of the novel SARS-CoV-2 coronavirus and the global
  COVID-19 pandemic has led to explosive growth in scientific research.
  Alas, much of the research in the literature lacks conditions to be
  reproducible, and recent publications on the association between
  population density and the basic reproductive number of SARS-CoV-2 are
  no exception. Relatively few papers share code and data sufficiently,
  which hinders not only verification but additional experimentation. In
  this paper, an example of reproducible research shows the potential of
  spatial analysis for epidemiology research during COVID-19.
  Transparency and openness means that independent researchers can, with
  relatively modest efforts, verify findings and use different
  approaches as appropriate. Given the high stakes of the situation, it
  is essential that scientific findings, on which good policy depends,
  are as robust as possible; as the empirical example shows,
  reproducibility is one of the keys to ensure this.
  \end{abstract}

@@

Given the ruinous human and economic cost of the pandemic, there has
been a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has termed the
deluge of COVID-19 publications a ``carnage of substandard research''
(Bramstedt, 2020). Part of the challenge of maintaining quality
standards in published research is that, despite an abundance of
recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
practice reproducibility has remained a lofty and somewhat aspirational
goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
et al., 2018), and the situation does not appear to have changed
substantially since (Gustot, 2020; Sumner et al., 2020).

@@

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic (Florida et al., 2020); the implications of the pandemic for
urban planning, design, and management are the topic of ongoing research
(Sharifi and Khavarian-Garmsir, 2020). The fact that the worst of the
pandemic was initially felt in dense population centers such as Wuhan,
Milan, Madrid, and New York, unleashed a torrent of research into the
associations between density and the spread of the pandemic. he answers
to some important questions hang on the results of these research
efforts. For example, are lower density regions safer from the pandemic?
Are de-densification policies warranted, even if just in the short term?
And in the longer term, will the risks of life in high density regions
presage a flight from cities? What are the implications of the pandemic
for future urban planning and practice? Over the past year, numerous
papers have sought to throw light on the underlying issue of density and
the pandemic; nonetheless the results, as will be detailed next, remain
mixed. Further, to complicate matters, precious few of these studies
appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. To this end, a recent study by Sy et al. (2021) is chosen as an
example of reproducible research. The objective is not to malign the
analysis of these researchers, but rather to demonstrate the value of
openness to allow for independent verification and further analysis.
Open data and open code mean that an independent researcher can, with
only modest efforts, not only verify the findings reported, but also
examine the same data from a perspective which may not have been
available to the original researchers due to differences in disciplinary
perspectives, methodological traditions, and/or training, among other
possible factors. The example, which shows consequential changes in the
conclusions reached by different analyses, should serve as a call to
researchers to redouble their efforts to increase transparency and
reproducibility in research. The present paper, in addition, aims to
show how data can be packaged in well-documented, shareable units, and
code can be embedded into self-contained documents suitable for review
and independent verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, will be available in a public
repository\footnote{For peer-review purposes, the data package and code
  are currently in an anonymous Drive folder:
  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.

@@

As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
mathematical models of contagion are valid at small-to-medium spatial
scales (and presumably, small temporal scales too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and different time scales. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers. On the other hand, a person can
choose to (or be forced to as a matter of policy) not go to a restaurant
in the first place. Nonetheless, the idea that high density correlates
with high transmission is so seemingly sensible that it is often taken
for granted even at larger scales (e.g., Cruz et al., 2020; Micallef et
al., 2020). At larger scales, however, there exists the possibility of
behavioral adaptations, which are difficult to capture in the
mechanistic framework of differential equations (or can be missing in
agent-based models, e.g., Gomez et al., 2021); these adaptations, in
fact, can be a key aspect of disease transmission.

A plausible behavioral adaptation during a pandemic, especially one
broadcast as widely and intensely as COVID-19, is risk compensation.
Risk compensation is a process whereby people adjust their behavior in
response to their \emph{perception} of risk (Noland, 1995; Phillips et
al., 2011; Richens et al., 2000). In the case of COVID-19, Chauhan et
al. (Chauhan et al., 2021) have found that perception of risks in the US
varies between rural, suburban, and urban residents, with rural
residents in general expressing less concern about the virus. It is
possible that people who listened to the message of leaders saying that
they were safe from the virus because of low density may not have taken
adequate precautions. Conversely, people in dense places who could more
directly observe the impact of the pandemic may have become overly
cautious. Both Paez et al. (2020) and Hamidi et al. (2020b) posit this
mechanism (i.e., greater compliance with social distancing in denser
regions) to explain the results of their analyses. The evidence
available does indeed show that there were important changes in behavior
with respect to mobility during the pandemic (Harris and Branion-Calles,
2021; Jamal and Paez, 2020; Molloy et al., 2020); furthermore, shelter
in place orders may have had greater buy-in from the public in higher
density regions (Feyman et al., 2020; Hamidi and Zandiatashbar, 2021),
and the associated behavior may have persisted beyond the duration of
official social-distancing policies (Praharaj et al., 2020). In
addition, there is evidence that changes in mobility correlated with the
trajectory of the pandemic (Noland, 2021; Paez, 2020). Given the
potential for behavioral adaptation, the question of density becomes
more nuanced: it is not just a matter of proximity, but also of human
behavior, which is better studied using population-level data and
models.

@@

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. (2020) who did not find evidence of
significant correlation between population density and confirmed number
of cases per day \emph{in conditions of lockdown} in China. This finding
echoes the results of Paez et al. (2020), who in their study of
provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, (2020) found zero or negative
associations between population density and infection numbers/deaths by
country. Fielding-Miller et al. (2020) contrast their finding about
rural counties with a negative relationship between COVID-19 deaths and
population density in urban counties in the US. For their part, in their
investigation of doubling time, White and Hébert-Dufresne (2020)
identified a negative and significant correlation between population
density and doubling time in US states. Likewise, (2021) found a small
negative (and significant) association between population density and
COVID-19 morbidity in districts in Tehran. Finally, two of the most
complete studies in the US (by Hamidi et al. 2020a, and 2020b) used an
extensive set of controls to find negative and significant correlations
between density and COVID-19 cases and fatalities at the level of
counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial analysis.

@@

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. For their multivariate
analysis, SWN used mixed linear models. This is a reasonable modelling
choice: \(R_0\) is an interval-ratio variable that is suitably modeled
using linear regression; further, as SWN note there is a likelihood that
the process in not independent ``among counties within each state,
potentially due to variable resource allocation and differing health
systems across states'' (p.~3). A mixed linear model accounts for this
by introducing random components; in the case of SWN, these are random
intercepts at the state level. SWN estimated various models with
different combinations of variables, including median household income
and prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SWN concluded that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

One salient aspect of the analysis in SWN is that the basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made modelling decisions, in this case basing their
analysis only on counties with valid observations. A modeler with
expertise in spatial analysis would likely ask some of the following
questions on reading SWN's paper: how were missing counties treated?
What are the implications of the spatial sampling framework used in the
analysis? Is it possible to spatially interpolate the missing
observations? Was there spatial residual autocorrelation in the models,
or was the use of mixed models sufficient to capture spatial
dependencies? These questions are relevant and their implications
important. Fortunately, SWN are an example of a reasonably open,
reproducible research product: their paper is accompanied by (most of)
the data and (most of) the code used in the analysis. This means that an
independent expert can, with only a moderate investment of time and
effort, reproduce the results in the paper, as well as ask additional
questions.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
2020). But in many other cases the data are publicly available, which in
fact has commonly been the case with population-level COVID-19
information. Typically the provenance of the data is documented, but in
numerous studies the data themselves are not shared (Amadu et al., 2021;
Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
whether a graduate student or a seasoned scientist, collecting,
organizing, and preparing data for a project can take a substantial
amount of time. Pointing to the sources of data, even when these sources
are public, is a small step towards reproducibility-but only a very
small one. Faced with the prospect of having to recreate a data set from
raw sources is probably sufficient to dissuade all but the most
dedicated (or stubborn) researcher from independent verification. This
is true even if part of the data are shared (e.g., Wong and Li, 2020).
In other cases, data are shared, but the processes followed in the
preparation of the data are not fully documented (Ahmad et al., 2020;
Skórka et al., 2020). These processes matter, as shown by the errors in
the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis (Noury et al., 2021; Pequeno et
al., 2020; Wang et al., 2021). Making code available only ``on demand''
(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
journals offer the facility to share supplemental materials online. Then
there are those papers that more closely comply with reproducibility
standards, and share well-documented processes and data, as well as the
code used in any analyses reported (Feyman et al., 2020; Paez et al.,
2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
2020). Even in this case, the pressure to publish ``new findings''
instead of replication studies can act as a deterrent, perhaps
particularly for younger researchers\footnote{The present paper was desk
  rejected by three journals that had previously published research on
  population density and the spread of COVID-19; in one case, the paper
  was too opinionated for the journal, in the other two cases, the paper
  was not a ``good fit'' despite dealing with a nearly identical issue
  as papers previously published in said journals. This does not inspire
  much confidence in the commitment of journals to reproducibility in
  research.}.

In the following sections, the analysis of RWN is reproduced, some
relevant questions from the perspective of an independent researcher
with expertise in spatial analysis are asked, and the data are
reanalyzed.

@@

Table \ref{tab:swn-results} reproduces the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SWN). It is possible to verify that the results match, with only the
minor (and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is changed to one percent units, instead of the
ten percent units used by SWN). The mixed linear model gives random
intercepts (i.e., the intercept is a random variable), and the standard
deviation is reported in the fourth row of Table \ref{tab:swn-results}.
It is useful to map the random intercepts: as seen in Figure
\ref{fig:random-terms-map}, other things being equal, counties in Texas
tend to have somewhat lower values of \(R_0\) (i.e., a negative random
intercept), whereas counties in South Dakota tend to have higher values
of \(R_0\). The key of the analysis, after extensive sensitivity
analysis, is a robust finding that population density has a positive
association with the basic reproductive number. But does it?

@@

The preceding section shows that thanks to the availability of code and
data, it is possible to verify the results reported by SWN. As noted
earlier, though, an independent researcher might have wondered about the
implications of the spatial sampling procedure used by SWN. The decision
to use a sample of counties with reliable basic reproductive numbers,
although apparently sensible, results in a non-random spatial sampling
scheme. Turning our attention back to Figure \ref{fig:R0-map}, we form
the impression that many counties without reliable values of \(R_0\) are
in more rural, less dense parts of the United States. This impression is
reinforced when we overlay the boundaries of urban areas with population
greater than 50,000 on the counties with valid values of \(R_0\) (see
Figure \ref{fig:urban-areas-map}). The fact that \(R_0\) could not be
accurately computed in many counties without large urban areas does not
mean that there was no transmission of the virus: it simply means that
we do not know with sufficient precision to what extent that was the
case. The low number of cases may be related to low population and/or
low population density. This is intriguing, to say the least: by
excluding cases based on the ability to calculate \(R_0\) we are
potentially \emph{selecting} the sample in a non-random way.

@@

A problematic issue with non-random sample selection is that parameter
estimates can become unreliable, and numerous techniques have been
developed to address this. A model useful for sample selection problems
is Heckman's selection model (see Maddala, 1983). The selection model is
in fact a system of two equations, as follows: \[
\begin{array}{c}
y_i^{S*} = \beta^{S\prime}x_i^S+\epsilon_i^S\\
y_i^{O*} = \beta^{O\prime}x_i^O+\epsilon_i^O
\end{array}
\] \noindent where \(y_i^{S*}\) is a latent variable for the sample
selection process, and \(y_i^{O*}\) is the latent outcome. Vectors
\(x_i^S\) and \(x_i^O\) are explanatory variables (with the possibility
that \(x_i^S = x_i^S\)). Both equations include random terms (i.e.,
\(\epsilon_i^S\) and \(\epsilon_i^O\)). The first equation is designed
to model the \emph{probability} of sampling, and the second equation the
outcome of interest (say \(R_0\)). The random terms are jointly
distributed and correlated with parameter \(\rho\).

@@

A sample selection model is estimated using the same selection of
variables as SWN Model 3. This is Sample Selection Model 1 in Table
\ref{tab:selection-results}. The first thing to notice about this model
is that the sample selection process and the outcome are correlated
(\(\rho\ne0\) with 5\% of confidence). The selection equation indicates
that the probability of a county to be in the sample increases with
population density (but at a decreasing rate due to the
log-transformation), when travel by private modes is more prevalent, and
as median household income in the county is higher. This is in line with
the impression made by Figure \ref{fig:urban-areas-map} that counties
with reliable values of \(R_0\) tended to be those with larger urban
centers. Once that the selection probabilities are accounted for in the
model, several things happen with the outcomes model. First, the
coefficient for population density is still positive, but the magnitude
changes: in effect, it appears that the effect of density is more
pronounced than what SWN Model 3 indicated. The coefficient for percent
of private transportation changes signs. And the coefficient for median
household income is now significant.
